---
title: "Microcosm data analysis with ZIPLN"
format: 
  html:
    embed-resources: true
execute: 
  warning: false
  cache: true
---

## Useful libraries & functions

```{r packages, message=FALSE}
library(PLNmodels)
library(tidyverse)
library(gt)
library(viridis)
library(lemon)
```

## Data loading and basic pretratments

```{r data}
data(microcosm)
```

## Fit various PLN model 

Assess groups effects in PLN models.

```{r microcosm-PLN, echo = TRUE, message=FALSE, results='hide'}
PLN           <- PLN(Abundance ~ 1             + offset(log(Offset)), data = microcosm)
PLN_site      <- PLN(Abundance ~ 0 + site      + offset(log(Offset)), data = microcosm)
PLN_time      <- PLN(Abundance ~ 0 + time      + offset(log(Offset)), data = microcosm)
PLN_site_time <- PLN(Abundance ~ 0 + site_time + offset(log(Offset)), data = microcosm)
```

## Fit various ZI-PLN models

#### Playing with the ZI component

Try various way of modeling probability of being zero:

```{r microcosm-ZI, echo = TRUE, message=FALSE, results='hide'}
ZI            <- ZIPLN(Abundance ~ 1 + offset(log(Offset)) | 1       , data = microcosm)
ZI_site       <- ZIPLN(Abundance ~ 1 + offset(log(Offset)) | 0 + site, data = microcosm)
ZI_time       <- ZIPLN(Abundance ~ 1 + offset(log(Offset)) | 0 + time, data = microcosm)
ZI_site_time  <- ZIPLN(Abundance ~ 1 + offset(log(Offset)) | 0 + site_time,
                       data = microcosm)
```

#### Include covariates both in PLN and ZI components

```{r microcosm-ZIPLN, echo = TRUE, message=FALSE, results='hide'}
ZI_site_PLN_site <- ZIPLN(Abundance ~ 0 + site      + offset(log(Offset)) | 0 + site,
                          data = microcosm)
ZI_time_PLN_time <- ZIPLN(Abundance ~ 0 + time      + offset(log(Offset)) | 0 + time,
                          data = microcosm)
ZI_PLN_site_time <- ZIPLN(Abundance ~ 0 + site_time + offset(log(Offset)) | 0 + site_time,
                          data = microcosm)
```

Other possible combinations:

```{r microcosm-ZIPLN-more, echo = TRUE, message=FALSE, results='hide'}
ZI_site_PLN_site_time <- ZIPLN(Abundance ~ 0 + site_time + offset(log(Offset)) | 0 + site,
                               data = microcosm)
ZI_time_PLN_site_time <- ZIPLN(Abundance ~ 0 + site_time + offset(log(Offset)) | 0 + time,
                               data = microcosm)
ZI_site_time_PLN_site <- ZIPLN(Abundance ~ 0 + site + offset(log(Offset)) | 0 + site_time,
                               data = microcosm)
ZI_site_time_PLN_time <- ZIPLN(Abundance ~ 0 + time + offset(log(Offset)) | 0 + site_time,
                               data = microcosm)
```

## Locate the best model

```{r microcosm-ZIPLN-criteria, echo = FALSE}
rbind(
  PLN$criteria, PLN_time$criteria, PLN_site$criteria, PLN_site_time$criteria,
  ZI$criteria, ZI_time$criteria, ZI_site$criteria, ZI_site_time$criteria,
  ZI_time_PLN_time$criteria, ZI_site_PLN_site$criteria,
  ZI_time_PLN_site_time$criteria, ZI_site_PLN_site_time$criteria,
  ZI_site_time_PLN_time$criteria, ZI_site_time_PLN_site$criteria,
  ZI_PLN_site_time$criteria
) %>%
  as.data.frame(row.names = 
    c("PLN", "PLN time", "PLN site", "PLN site * time",
      "ZI", "ZI time", "ZI site", "ZI site * time",
      "ZI time PLN time", "ZI site PLN site",
      "ZI time and PLN site * time", "ZI site and PLN site * time",
      "ZI site * time and PLN time", "ZI site * time and PLN site",
      "ZI and PLN site * time")) %>%
  gt::gt(rownames_to_stub = TRUE) %>% 
  gt::tab_stubhead(label = "model") %>% 
  gt::tab_header(
    title = md("**Model selection criteria**"),
    subtitle = md("The *top two* largest are presented")
    ) %>% 
  gt::tab_row_group(
    label = "ZIPLN",
    rows = 9:15
  ) %>% 
  gt::tab_row_group(
    label = "ZI",
    rows = 5:8
  )  %>% 
  gt::tab_row_group(
    label = "PLN",
    rows = 1:4
  ) %>% 
  gt::tab_options(table.font.size = 18) %>% 
      tab_style(
        style = list(
          cell_fill(color = '#9a2515'),
          cell_text(color = 'white')
          ),
          locations = cells_body(
          columns = c("BIC"), 
          rows = c(1, 7, 10)
    )) %>% 
      tab_style(
        style = list(
          cell_fill(color = '#9a2515'),
          cell_text(color = 'white')
          ),
          locations = cells_body(
          columns = c("AIC"), 
          rows = c(4, 8, 14)
    )) %>% 
      tab_style(
        style = list(
          cell_fill(color = '#9a2515'),
          cell_text(color = 'white')
          ),
          locations = cells_body(
          columns = c("ICL"), 
          rows = c(2, 8, 15)
    ))
```

$\rightsquigarrow$ Ok, for illustrative purpose, let us keep model with `site_time` in both ZI an PLN components.

```{r best-models}
best_PLN   <- PLN_site_time
best_ZIPLN <- ZI_PLN_site_time
```

## Model fits 

### ZI-PLN

```{r zipln-fit}
latent_means <- ZI_site_time_PLN_site$latent
colnames(latent_means) <- colnames(microcosm$Abundance)
asv_clust <- hclust(dist(t(latent_means)), method = "ward.D2")
asv_ordering <- asv_clust$labels[asv_clust$order]

plot_data <- latent_means |> 
  as_tibble(rownames = "sample") |> mutate(sample = microcosm$sample) %>% 
  pivot_longer(-sample, names_to = "ASV", values_to = "M", names_ptypes = factor(levels = asv_ordering)) |> 
  inner_join(microcosm, by = "sample")

proba_zi <- ZI_site_time_PLN_site$var_par$R
colnames(proba_zi) <- colnames(microcosm$Abundance)
plot_data_zi <- proba_zi |> 
  as_tibble(rownames = "sample") |> mutate(sample = microcosm$sample) %>% 
  pivot_longer(-sample, names_to = "ASV", values_to = "M", names_ptypes = factor(levels = asv_ordering)) |> 
  inner_join(microcosm, by = "sample")

sigma <- sigma(ZI_site_time_PLN_site)
counts_tibble <- microcosm$Abundance |> as("matrix") |> 
  as_tibble(rownames = "sample") |> 
  pivot_longer(-sample, names_to = "ASV", values_to = "count", names_ptypes = factor(levels = asv_ordering))

p <- inner_join(plot_data, counts_tibble, by = join_by(sample, ASV)) |> 
  inner_join(plot_data_zi |> select(sample, ASV, ZI = M), by = join_by(sample, ASV)) |> 
  inner_join(tibble(ASV = factor(asv_ordering, levels = asv_ordering), 
                    S2  = diag(sigma)[asv_ordering]), 
             by = join_by(ASV)) |> 
  ggplot(aes(x = count, y = (1 - ZI)*exp(M + 0.5*S2))) + 
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, color = "grey60", linetype = "dashed") + 
  scale_x_log10() + scale_y_log10() + 
  labs(x = "Observed", y = "Fitted") + 
  facet_wrap(~site_time) + 
  theme_bw()
p
```

## Model fits

### PLN seems to do just as well...

```{r pln-fit}
latent_means_PLN <- best_PLN$latent
colnames(latent_means_PLN) <- colnames(microcosm$Abundance)
sigma_PLN <- sigma(best_PLN)

plot_data_PLN <- latent_means_PLN |> 
  as_tibble(rownames = "sample") |> mutate(sample = microcosm$sample) %>% 
  pivot_longer(-sample, names_to = "ASV", values_to = "M", names_ptypes = factor(levels = asv_ordering)) |> 
  inner_join(microcosm, by = "sample")

counts_tibble <- microcosm$Abundance |> as("matrix") |> 
  as_tibble(rownames = "sample") |> 
  pivot_longer(-sample, names_to = "ASV", values_to = "count", names_ptypes = factor(levels = asv_ordering))
p <- inner_join(plot_data_PLN, counts_tibble, by = join_by(sample, ASV)) |> 
  inner_join(tibble(ASV = factor(asv_ordering, levels = asv_ordering), 
                    S2  = diag(sigma)[asv_ordering]), 
             by = join_by(ASV)) |> 
  ggplot(aes(x = count, y = exp(M + 0.5*S2))) + 
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, color = "grey60", linetype = "dashed") + 
  scale_x_log10() + scale_y_log10() + 
  labs(x = "Observed", y = "Fitted") + 
  facet_wrap(~site_time) + 
  theme_bw()
p
```

## Fits of zeros

### But zeros are not well predicted 

```{r fit-zeros}
# distribution of zeros
true_zeros <- microcosm$Abundance == 0
n_zeros <- sum(true_zeros)
n_nzeros <- sum(!true_zeros)
distr_zeros <-
    data.frame(
        fitted   = c(as.vector(fitted(best_PLN)[true_zeros]), as.vector(fitted(ZI_site_time_PLN_site)[true_zeros])),
        obs      = rep(1:n_zeros, 2),
        method   = factor(rep(c("PLN site_time", "ZI site_time"), each = n_zeros))
    ) %>%
    ggplot() +
    geom_rug(aes(x = fitted, color = method)) +
    geom_point(aes(x = obs, y = fitted, color = method), size = 0.3, alpha = 0.1) +
    ggtitle("Distribution of observed zeros as fitted by the models") +
    xlab(label = "observations #") +
    scale_color_viridis_d() + theme_bw()
distr_zeros
```

## ZI-PLN: latent layer

### Latent means of the PLN component (w/wo covariate effects)

```{r palette, include = FALSE}
library(ggtext)
library(factoextra)
library(ggh4x)
library(patchwork)
site_time_palette <- c(
  "_**<span style = 'color:#6baed6;'>Buccal</span>**_"  = "transparent",
  "B -1W" = "#bdd7e7", 
  "B 1M" = "#6baed6", 
  "B 3M" = "#2171b5", 
  "B 7M" = "#053061",
  "_**<span style = 'color:#fb6a4a;'>Nasal</span>**_"  = "transparent",
  "N -1W" = "#fcae91", 
  "N 1M" = "#fb6a4a", 
  "N 3M" = "#cb181d",
  "N 7M" = "#67001F",
  "_**<span style = 'color:#bf812d;'>Vaginal</span>**_"  = "transparent",
  "V -1W" = "#dfc27d", 
  "V 1M" = "#bf812d", 
  "V 3M" = "#8c510a",
  "V 7M" = "#543005",
  "_**<span style = 'color:#74c476;'>Milk</span>**_"  = "transparent",
  "L -1W" = "#bae4b3", 
  "L 1M" = "#74c476", 
  "L 3M" = "#238b45", 
  "L 7M" = "#00441B"
)
site_time_palette_simple <- site_time_palette[-c(1,6,11,16)]
index_palette <- c("minus" = "#67a9cf", 
                   "plus" = "#ef8a62")
site_time_design <- strip_nested(
  background_x = elem_list_rect(fill = c(
    ## Time
    rep("gray", 4), 
    ## Site within time
    site_time_palette_simple[c(1, 5, 9, 2, 6, 10, 14, 3, 7, 11, 15, 4, 8, 12, 16)]))
)
```

```{r pca_zipln, fig.height=5, fig.width=10}
pca_PLN_latent_means <- prcomp(best_ZIPLN$latent) %>% factoextra::fviz_pca_ind(axes = c(1,2), col.ind = microcosm$site_time, palette = site_time_palette)
pca_PLN_latent_means_wcov <- prcomp(best_ZIPLN$latent_pos) %>% factoextra::fviz_pca_ind(axes = c(1,2), col.ind = microcosm$site_time, palette = site_time_palette)
pca_PLN_latent_means + pca_PLN_latent_means_wcov + plot_annotation(
  title = "PCA on latent means", 
  subtitle = "With (left: M) or w/o (right: M - XB) the effect of covariates"
)
```

## PLN: latent layer

### Same for standard PLN

```{r pca_pln, fig.height=5, fig.width=10}
pca_PLN_latent_means <- prcomp(best_PLN$latent) %>% factoextra::fviz_pca_ind(axes = c(1,2), col.ind = microcosm$site_time, palette = site_time_palette)
pca_PLN_latent_means_wcov <- prcomp(best_PLN$latent_pos) %>% factoextra::fviz_pca_ind(axes = c(1,2), col.ind = microcosm$site_time, palette = site_time_palette)
pca_PLN_latent_means + pca_PLN_latent_means_wcov + plot_annotation(
  title = "PCA on latent means", 
  subtitle = "With (left: M) or w/o (right: M - XB) the effect of covariates"
)
```

## ZI PLN

### Clustering of the latent means vs probability of zero inflation

```{r heatmap_zipln, fig.height=5, fig.width=10}
latent_means <- best_ZIPLN$latent
colnames(latent_means) <- colnames(microcosm$Abundance)
asv_clust <- hclust(dist(t(latent_means)), method = "ward.D2")
asv_ordering <- asv_clust$labels[asv_clust$order]

plot_data <- latent_means |> 
  as_tibble(rownames = "sample") |> mutate(sample = microcosm$sample) %>% 
  pivot_longer(-sample, names_to = "ASV", values_to = "M", names_ptypes = factor(levels = asv_ordering)) |> 
  inner_join(microcosm, by = "sample")
p_heatmap <- plot_data |> 
  ggplot(aes(x = sample, y = ASV, fill = M)) + 
  geom_tile() + 
  # scale_fill_distiller(type = "div", ) +
  scale_fill_viridis_c(option = "E") + 
  facet_nested(~ time + site, scales = "free_x", space = "free_x", strip = site_time_design) + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.spacing.x = unit(0.1, "line")) +
  labs(x = "Samples", y = "ASV") + 
  NULL

proba_zi <- best_ZIPLN$var_par$R
colnames(proba_zi) <- colnames(microcosm$Abundance)
plot_data_zi <- proba_zi |> 
  as_tibble(rownames = "sample") |> mutate(sample = microcosm$sample) %>% 
  pivot_longer(-sample, names_to = "ASV", values_to = "M", names_ptypes = factor(levels = asv_ordering)) |> 
  inner_join(microcosm, by = "sample")

p_heatmap_zi <- plot_data_zi |> 
  ggplot(aes(x = sample, y = ASV, fill = M)) + 
  geom_tile() + 
  # scale_fill_distiller(type = "div", ) +
  scale_fill_viridis_c(option = "E", name = "P(Y = 0)") + 
  facet_nested(~ time + site, scales = "free_x", space = "free_x", strip = site_time_design) + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.spacing.x = unit(0.1, "line")) +
  labs(x = "Samples", y = "ASV") + 
  NULL

(p_heatmap + p_heatmap_zi & theme(legend.position = "bottom")) + 
  plot_layout() + 
  plot_annotation(
    title = "Heatmap of latent means (M, left) and ZI probability (Rho, right)"
  )
```


## ZI PLN

### Latent means vs probability of zero inflation

```{r means_vs_zi}
plot_data |> 
  select(ASV, sample, M, site_time) |> 
  bind_cols(plot_data_zi |> select(ZI = M)) |>  
  ggplot(aes(x = M, y = ZI, color = site_time)) + 
  geom_point(alpha = 0.1) + 
  facet_wrap(~site_time) + 
  scale_color_manual(values = site_time_palette, guide = "none") + 
  # scale_y_log10() + 
  labs(x = "Latent mean", y = "Proba of zero inflation") + 
  NULL
```

## Residual covariance

### Reordered by hierarchical clustering with complete linkage

```{r zipln-covariance}
sigma_PLN <- sigma(best_PLN)
par(mfrow = c(2,2))
sigma |> cov2cor() %>% corrplot::corrplot(method = "color", tl.pos = "n", order = "hclust", is.corr = TRUE, hclust.method = "complete", title = "\nCorrelation matrix ZIPLN")
sigma_PLN |> cov2cor() %>%  corrplot::corrplot(method = "color", tl.pos = "n", order = "hclust", is.corr = TRUE, hclust.method = "complete", title = "\nCorrelation matrix PLN")
hist(sigma, main = "histogramm of ZIPLN covariances")
hist(sigma_PLN, main = "histogramm of PLN covariances")
```

