---
title: "Revisiting PLNmodels output with htmlwidgets"
subtitle: "The Oaks Powdery Mildew Data Set"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  screenshot.force = FALSE, 
  echo = TRUE,
  rows.print = 5,
  fig.width = 10, 
  fig.height = 10,
  message = FALSE, 
  warning = FALSE)
```

## Preliminaries

The packages required for the analysis are **PLNmodels** plus some others for data manipulation and representation:

```{r dependencies}
library(PLNmodels)
library(tidyverse)
library(ggplot2)
nb_cores <- 4
```

Here are a collection of packages implementing various JS widgets:

```{r widgets}
library(DT)
library(scatterD3)
library(plotly)
library(edgebundleR)
library(threejs)
library(networkD3)
```

The _oaks amplicon data set_ at play gives the abundance of 114 taxa (66 bacterial OTU, 48 fungal OTUs) in 116 samples. For each sample, 11 additional covariates are known

```{r loading}
data(oaks)
datatable(oaks$Abundance[, 1:5] )
```

## PLNmodels Analyses 

### PCA analysis

```{r principal component analysis, cache = TRUE, warning = FALSE, results='hide'}
my_PCAs <- PLNPCA(Abundance ~ 1 + offset(log(Offset)), data = oaks, ranks = 25:30, control_main = list(cores = nb_cores))
```

### Sparse precision matrix (Network)

```{r network analysis, cache = TRUE, results='hide'}
my_networks <- PLNnetwork(Abundance ~ 0 + tree + offset(log(Offset)), data = oaks, control_main = list(trace = 2))
```

## Outputs and Vizualisation

### Principal Components Map 

We explore various scatterplot solutions to represent the individual factor map of the PLN PCA.

```{r data for scatterplot}
coord <- my_PCAs$getBestModel()$scores[, 1:3] %>% 
  as.data.frame() %>%   
  setNames(c("PC1", "PC2", "PC3")) %>% 
  add_column(tree = oaks$tree, names = rownames(oaks$Abundance))
```

#### Direct "plotlyfication"

```{r biplot factoextra1}
my_PCA <- my_PCAs$getBestModel()
t(tcrossprod(my_PCA$model_par$B, my_PCA$var_par$M)) %>%
  prcomp(center = FALSE, scale. = FALSE) %>%
  factoextra::fviz_pca_biplot(select.var = list(contrib = 10), col.ind  = oaks$tree,
                              title = "Biplot after correction (10 most contributing species, samples colored by distance to ground)") +
  labs(col = "distance (cm)") + scale_color_viridis_d()
```

```{r factoextra}
ggplotly()
```

#### ScatterD3 (Another fancy scatterplot)

```{r scatterD3}
scatterD3(data = coord, x = PC1, y = PC2, lab = names,
          col_var = tree, symbol_var = tree,
          xlab = "PC1", ylab = "PC2", col_lab = "tree",
          symbol_lab = "tree", lasso = TRUE)
```

#### Native plotly (3D scatterplot)

```{r plotly}
fig <- plot_ly(
  coord, x = ~PC1, y = ~PC2, z = ~PC3, color = ~tree, size = .35,
  text = ~paste('status:', tree), type = "scatter3d") %>% 
  layout(title = "Individual Factor Map of the Oaks powdery Mildew data set",
         scene = list(xaxis = list(title = 'PC1'),
                      yaxis = list(title = 'PC2'),
                      zaxis = list(title = 'PC3'))
  )
fig
```

#### threejs (3D scatterplot)

```{r threejs}
group <- rainbow(3)[as.numeric(oaks$tree)]
coord %>% select(1:3) %>% as.matrix() %>% 
scatterplot3js(col = group, size = 0.25, pch = ".", grid = FALSE, bg = "black")
```

### Networks

```{r net selection}
my_net <- my_networks$getBestModel('EBIC')
```

#### EdgebundleR


```{r edgbundle, results='asis', fig.show='asis'}
g <- my_net$plot_network(output = "igraph", plot = FALSE)
edgebundle(g)
```

#### networkD3

```{r networkD3, results='asis', fig.show='asis'}
# Convert to object suitable for networkD3
d3 <- igraph_to_networkD3(g, group = sapply(strsplit(colnames(oaks$Abundance), "_"), function(x) x[[1]]))

# Create force directed network plot
forceNetwork(Links = d3$links, Nodes = d3$nodes, 
             Source = 'source', Target = 'target', 
             NodeID = 'name', Group = 'group', opacity = 0.4)
```

