---
title: "PLNmodels: Poisson lognormal models for multivariate count data"
format:
  gfm:
    default-image-extension: ""
bibliography: vignettes/article/PLNreferences.bib
link-citations: true
---

<!-- badges: start -->

[![R-CMD-check](https://github.com/PLN-team/PLNmodels/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PLN-team/PLNmodels/actions/workflows/R-CMD-check.yaml) [![Coverage status](https://codecov.io/gh/pln-team/PLNmodels/branch/master/graph/badge.svg)](https://codecov.io/github/pln-team/PLNmodels?branch=master) [![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/PLNmodels)](https://cran.r-project.org/package=PLNmodels) [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html) [![](https://img.shields.io/github/last-commit/pln-team/PLNmodels.svg)](https://github.com/pln-team/PLNmodels/commits/master)

<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  cache      = TRUE,
  cache.lazy = FALSE, # R6 model objects don't survive knitr's lazy-load cache promises
  fig.path   = "man/figures/README-",
  fig.width  = 7, fig.height = 5
)
```

## Description

> The Poisson lognormal model and variants [@chiquet2021] can be used for a variety of multivariate problems when count data are at play. This package implements efficient variational algorithms to fit such models, accompanied with a set of functions for visualization and diagnostic. See [all the dedicated vignettes](https://pln-team.github.io/PLNmodels/articles/) for a comprehensive introduction.

**PLNmodels** covers the following models, all built around the multivariate Poisson-lognormal distribution and sharing a common formula-based interface (covariates, offsets, weights) and a choice of optimization backends (a fast built-in Newton solver, NLOPT, and an experimental torch backend):

- **PLN** [@AiH89]: unpenalized multivariate Poisson regression, with several covariance structures (full, diagonal, spherical, fixed, or a genetic/heritability structure).
- **PLNPCA** [@PLNPCA]: probabilistic Poisson PCA — a rank-constrained covariance for dimension reduction and visualization.
- **PLNLDA**: Poisson lognormal discriminant analysis [@fisher1936] for the supervised classification of count data.
- **PLNnetwork** [@PLNnetwork]: sparse inverse-covariance (network) inference via a graphical-lasso-like penalty [@FHT08].
- **PLNmixture**: model-based clustering [@fraley1999] of count data via a mixture of PLN models.
- **ZIPLN** [@ZIPLN]: a zero-inflated extension of PLN for data with excess zeros, with the same family of covariance structures and an optional sparse (`ZIPLNnetwork`) variant [@ZIPLNnetwork].

## Installation

**PLNmodels** is available on [CRAN](https://cran.r-project.org/package=PLNmodels). The development version is available on [GitHub](https://github.com/pln-team/PLNmodels).

```{r package CRAN, eval = FALSE}
install.packages("PLNmodels")             # last stable version, from CRAN
remotes::install_github("pln-team/PLNmodels")            # development version, from GitHub
remotes::install_github("pln-team/PLNmodels@tag_number")  # a specific tagged release
```

## Illustration

We illustrate the main models on the `barents` data set [@fossheim2006]: the abundance of 30 fish species observed in 89 sites in the Barents sea, along with depth, temperature and geographic coordinates for each site.

```{r load}
library(PLNmodels)
data(barents)
## a simple North/South split of the sites, used below to illustrate PLNLDA
barents$zone <- factor(ifelse(barents$Latitude > median(barents$Latitude), "North", "South"))
```

### PLN: fit and inspect the covariance structure

```{r pln}
myPLN <- PLN(Abundance ~ Depth + Temperature + offset(log(Offset)), data = barents)
myPLN
```

```{r pln-corrplot}
corrplot::corrplot(cov2cor(sigma(myPLN)), order = "AOE", type = "upper", tl.cex = 0.6)
```

### PLNLDA: discriminant analysis

```{r plnlda}
myLDA <- PLNLDA(Abundance ~ offset(log(Offset)), grouping = zone, data = barents)
plot(myLDA)
```

### PLNPCA: dimension reduction

```{r plnpca}
myPCAs <- PLNPCA(Abundance ~ Depth + Temperature + offset(log(Offset)), data = barents, ranks = 1:5)
myPCA  <- getBestModel(myPCAs)
factoextra::fviz_pca_biplot(
  myPCA, select.var = list(contrib = 10), col.ind = barents$Temperature,
  title = "Biplot (10 most contributing species, sites colored by temperature)"
) + ggplot2::labs(col = "Temperature") + ggplot2::scale_color_viridis_c()
```

### PLNnetwork: sparse network inference

```{r plnnetwork}
myNets <- PLNnetwork(Abundance ~ Depth + Temperature + offset(log(Offset)), data = barents)
plot(getBestModel(myNets), remove.isolated = TRUE)
```

### PLNmixture: model-based clustering

```{r plnmixture}
my_mixtures <- PLNmixture(Abundance ~ offset(log(Offset)), data = barents, clusters = 1:4,
                           control = PLNmixture_param(smoothing = "none"))
myMixture <- getBestModel(my_mixtures)
plot(myMixture, "pca", main = "Clustering membership in the individual factor map")
table(cluster = myMixture$memberships, zone = barents$zone)
```

## References
