---
title: "Analyzing multivariate count table with the Poisson log-normal model"
author: "PLN team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates the use of the `PLN` function and the methods accompaning the R6 class `PLNfit`. From the statistical point of view, the function `PLN` adjusts a multivariate Poisson lognormal model to a table of counts. `PLN` is the building block for all the multivariate models found in the `PLNmodels` package: having a basic understanding of both  the mathematical background and the associated set of `R` functions is good place to start. 

We will illustrate our point with an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence. 

```{r data_load}
library(PLNmodels)
data(trichoptera)
grouping <- trichoptera$Group
```

Mathematical background and a full description of the trichoptera data set can be found in [the introductory vignette](PLNmodels_introduction.html).

### A PLN model with latent main effects

The model can be fitted with the function `PLN` as follows:

```{r simple PLN}
myPLN <- PLN(Abundance ~ 1 + offset(log(TotalCounts)), trichoptera)
```

Note the use of the `formula` object to specify the model: the vector $\boldsymbol\mu$ of main effects in the mathematical formulation (one per column species) is specified in the call with the formula `~ 1`. We use the function offset with a log transform of the total count. Abundace is a variable in ther data frame `trichoptera` correspondings to a matrix of 17 columns.

`myPLN` is an `R6` object with class `PLNfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
myPLN
```

#### Field access to a PLN output

Let us try to access the most common fields:

```{r fields-access}
myPLN$loglik
myPLN$BIC
myPLN$ICL
myPLN$R_squared
myPLN$criteria
```

We can also reach the matrix of regression parameters and the variance/covariance matrix of the latent variable $Z$ with the traditional function found in `R` for (g)lm manipulation:

```{r coef vcov}
head(coef(myPLN))
head(vcov(myPLN))
```

Consider plotting the parameters with the nice `corrplot` package, _e.g._ the the covariance matrix:
```{r plot covariance, fig.width=7, fig.height=5}
corrplot::corrplot(vcov(myPLN), is.corr = FALSE)
```

#### Observation weights

It is also possible to use observation weights like in standard (generalized) linear models:

```{r weighted, fig.width=7, fig.height=5}
myPLN_weighted <- PLN(Abundance ~ 1 + offset(log(TotalCounts)), data = trichoptera, weights = runif(nrow(trichoptera)))
corrplot::corrplot(vcov(myPLN) - vcov(myPLN_weighted), is.corr = FALSE)
```

#### Make prediction with new data

The prediction method can be use to predict new responses from a new table of data (with both covariates and offset if needed.)

```{r prediction}
newdata <- trichoptera[1:10, ]
newdata$Abundance <- NULL
predict(myPLN, newdata = newdata, type = "response")
```

