---
title: "Analyze of the Trichopetra data set with the multivariate Poisson log-normal model"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates the use of the `PLN` function and the methods accompaning the R6 class `PLNfit`. From the statistical point of view, the function `PLN` adjusts a multivariate Poisson lognormal model to a table of counts. Such data are common in many application fields such as ecology (abundance table), or genomics (single cell data). These data often comes with some external knowledge (via covariates and offsets) which should be taken into account during the fits. 

`PLN` is the building block for all the multivariate models found in the `PLNmodels` package: having a basic understanding of both  the mathematical background and the associated set of `R` functions is good place to start. 

We will illustrate our point with an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence. 

## Mathematical background

The multivariate Poisson lognormal model relates some $p$-dimensional observation vectors $\mathbf{Y}_i$ to some  $p$-dimensional vectors of Gaussian latent variables $\mathbf{Z}_i$ as follows
\begin{equation}
  \label{eq:pln-model}
  \begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}({\boldsymbol\mu},\boldsymbol\Sigma) \\
  \text{observation space } &  Y_{ij} | Z_{ij} \quad \text{indep.} &   \mathbf{Y}_i | \mathbf{Z}_i\sim\mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right)
  \end{array}
\end{equation}

The parameter ${\boldsymbol\mu}$ corresponds to the main effects and the latent covariance matrix $\boldsymbol\Sigma$ describes the underlying structure of dependence between the $p$ variables. 

This model generalizes naturally to a formulation closer to a multivariate generalize linear model, where the main effect is due to a linear combination of $d$ covariates $\mathbf{x}_i$. We also let the possibility to add some offsets for the $d$ variables and in each sample., that is $\mathbf{o}_i$. Hence, model \eqref{eq:pln-model} generalizes to

\begin{equation}
  \label{eq:pln-model-glm}
  \mathbf{Y}_i | \mathbf{Z}_i \sim \mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right), \qquad \mathbf{Z}_i \sim \mathcal{N}({\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta},\boldsymbol\Sigma) \\
\end{equation}
where $\boldsymbol\Theta$ is a $d\times p$ matrix of regression parameters. At the end of the day, the data matrices available to feed the model are 

  - the $n\times p$ matrix of counts $\mathbf{Y}$ 
  - the $n\times p$ matrix of covariates $\mathbf{X}$ 
  - the $n\times p$ matrix of offsets $\mathbf{O}$ 

## Analysis of the Trichoptera data set

### Trichoptera data 

The trichoptera data set is included in the package:

```{r data_load}
library(PLNmodels)
data(trichoptera)
```

See `?PLNmodels::trichoptera` for details. 

The `data.frame` `trichoptera` includes 49 rows (the observations - or trapping nights) and 14 columns. 

```{r trichoptera description}
colnames(trichoptera)
```

The columns are all univariate covariates expect the `Abundance` column, which is a $49 \times 17$ matrix of abundancies for each 17 species during the 49 trapping nights. It corresponds to the matrix $\mathbf{Y}$ in our model: 

```{r responses_trichoptera}
trichoptera$Abundance[1:5, ]
```

It seems natural to use an offset term for each observation that corresponds to the total counts per night, in order to correct for the sampling effort. It can be specified by a $49 \times 17$ matrix of offsets (the matrix $\mathbf{O}$ in our model) 

```{r offset_trichopera}
trichoptera$TotalCounts[1:5, ]
```
Note that the offset term remains the same in a given sample albeit sometimes one might include an offset specific to both the sample and the species.

The matrix $49 \times 12$ entries remaings corresponds to the matrix $\mathbf{X}$ in our model:

```{r covariates_trichoptera}
head(trichoptera[, -c(1,14)])
```

### A model with latent main effects for the Trichopetra data set

The model can be fitted with the function `PLN` as follows:

```{r}
myPLN <- PLN(Abundance ~ 1 + offset(log(TotalCounts)), trichoptera)
```

Note the use of the `formula` object to specify the model: the vector $\boldsymbol\mu$ of main effects in the mathematical formulation (one per column species) is specified in the call with the formula `~ 1`. We use the function offset with a log transform of the total count. Abundace is a variable in ther data frame `trichoptera` correspondings to a matrix of 17 columns.

`myPLN` is an `R6` object with class `PLNfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
myPLN
```

#### Field access to a PLN output

Let us try to access the most common fields:

```{r fields-access}
myPLN$loglik
myPLN$BIC
myPLN$ICL
myPLN$R_squared
myPLN$criteria
```

We can also reach the matrix of regression parameters and the variance/covariance matrix of the latent variable $Z$ with the traditional function found in `R` for (g)lm manipulation:

```{r coef vcov}
head(coef(myPLN))
head(vcov(myPLN))
```

Consider plotting the parameters with the nice `corrplot` package, _e.g._ the the covariance matrix:
```{r plot covariance, fig.width=7, fig.height=5}
corrplot::corrplot(vcov(myPLN), is.corr = FALSE)
```

#### Observation weights

It is also possible to use observation weights like in standard (generalized) linear models:

```{r weighted, fig.width=7, fig.height=5}
myPLN_weighted <- PLN(Abundance ~ 1 + offset(log(TotalCounts)), data = trichoptera, weights = runif(nrow(trichoptera)))
corrplot::corrplot(vcov(myPLN) - vcov(myPLN_weighted), is.corr = FALSE)
```

#### Make prediction with new data

The prediction method can be use to predict new responses from a new table of data (with both covariates and offset if needed.)

```{r prediction}
newdata <- trichoptera[1:10, ]
newdata$Abundance <- NULL
predict(myPLN, newdata = newdata, type = "response")
```

