---
title: "Analyzing the Trichopetra data set with the Poisson log-normal model"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette intends to illustrate the basical use of the `PLN` function and the methods accompaning the R6 Classes `PLNfamily` and `PLNfit`. To this end, let us consider an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence. 

```{r data_load}
library(PLNmodels)
data(trichoptera)
```

See `?PLNmodels::trichoptera` for details. 

## Poisson log-normal models

The family of models available in the package belongs to the family of probabilistic PCA models, where the observation are not necessarily Gaussian.  In particular, we consider that observations are drawn from a Poisson lognormal distribution.

### A quick view of the mathematical formulation of the model

The Poisson lognormal model relates some $p$-dimensional observation vectors $\mathbf{Y}_i$ to some  $p$-dimensional vectors of Gaussian latent variables $\mathbf{Z}_i$ as follows
\begin{equation}
  \label{eq:pca-model}
  \begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}{\boldsymbol\mu},\boldsymbol\Sigma) \\
  \text{observation space } &  Y_{ij} | Z_{ij} \quad \text{indep.} & Y_{ij} | Z_{ij} \sim \mathcal{P}\left(\exp\{Z_{ij}\}\right)
  \end{array}
\end{equation}

The parameter ${\boldsymbol\mu}$ corresponds to the main effects and the latent covariance matrix $\boldsymbol\Sigma$ describes the underlying structure of dependence between the $p$ variables.

### A model with latent main effects for the Trichopetra data set

We start by adjusting the above model to Trichoptera data set. Here, the matrix of observations $\mathbf{Y}$ corresponds to the abundance of each species in each trapping night. 

```{r responses_trichoptera}
str(trichoptera$Abundance)
```


It seems also natural to add an offset corresponding to the total count per night.
```{r offset_trichopera}
TotalCount <- matrix(
  rowSums(trichoptera$Abundance),
  nrow = nrow(trichoptera$Abundance),
  ncol = ncol(trichoptera$Abundance)
)
```


The model can be fitted with the function `PLN` as follows:

```{r}
myPLN <- PLN(Abundance ~ 1 + offset(log(TotalCount)), trichoptera)
```

Note the use of the `formula` object to specify the model: the vector $\boldsymbol\mu$ of main effects in the mathematical formulation (one per column species) is specified in the call with the formula `~ 1`. We use the function offset with a log transform of the total count. Abundace is a variable in ther data frame `trichoptera` correspondings to a matrix of 17 columns.

`myPLN` is an `R6` object with class `PLNfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
myPLN
```

Let us try this a bit:

```{r fields-access}
myPLN$loglik
myPLN$BIC
myPLN$ICL
myPLN$R_squared
myPLN$criteria
```

Complementary information comes with the `plot` method:
```{r plot_model, fig.width=7, fig.height=5}
myPLN$plot("model")
```

```{r plot_var, fig.width=7, fig.height=5}
myPLN$plot("variational")
```

It is also possible to use observation weights like in standard (generalized) linear models:

```{r weighted, fig.width=7, fig.height=5}
myPLN_weighted <- PLN(trichoptera$Abundance ~ 1 + offset(log(TotalCount)), weights = runif(nrow(TotalCount)))
myPLN_weighted$plot("model")
myPLN_weighted$plot("variational")
```

