---
title: "Analyzing the Trichopetra data set with the Poisson log-normal model"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette intends to illustrate the basical use of the `PLN` function and the methods accompaning the R6 Classes `PLNfamily` and `PLNfit`. To this end, let us consider an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence.

```{r child = 'trichoptera.Rmd'}
```

```{r child = 'PLNmodels.Rmd'}
```

### A model with latent main effects for the Trichopetra data set

We start by adjusting the above model to Trichoptera data set. Here, the matrix of observations $\mathbf{Y}$ corresponds to the abundance of each species in each trapping night. The model can be fitted with the function `PLN` for a series of penalties contorling the number of edges in the network as follows:

```{r}
abundance <- as.matrix(trichometeo$fau) ## must be a matrix
model.nocov <- PLN(abundance ~ 1)
```

Note the use of the `formula` object to specify the model: the vector $\boldsymbol\mu$ of main effects in the mathematical formulation (one per column species) is specified in the call with the formula `~ 1`.

The `model.nocov` variable is an `R6` object with class `PLNfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
model.nocov
```

Let us try this a bit:

```{r fields-access}
model.nocov$loglik
model.nocov$BIC
model.nocov$ICL
model.nocov$R_squared
model.nocov$criteria
```

Complementary information comes with the `plot` method:
```{r plot_model, fig.width=7, fig.height=5}
model.nocov$plot("model")
```

```{r plot_var, fig.width=7, fig.height=5}
model.nocov$plot("variational")
```

