---
title: "Analyzing the Trichopetra data set with the Poisson log-normal model"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette intends to illustrate the basical use of the `PLN` function and the methods accompaning the R6 Classes `PLNfamily` and `PLNfit`. To this end, let us consider an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence. 

```{r data_load}
library(PLNmodels)
data(trichoptera)
```

See `?PLNmodels::trichoptera` for details. 

```{r child = 'PLNmodels.Rmd'}
```

### A model with latent main effects for the Trichopetra data set

We start by adjusting the above model to Trichoptera data set. Here, the matrix of observations $\mathbf{Y}$ corresponds to the abundance of each species in each trapping night. 

```{r responses_trichoptera}
str(trichoptera$Abundance)
```


It seems also natural to add an offset corresponding to the total count per night.
```{r offset_trichopera}
TotalCount <- matrix(
  rowSums(trichoptera$Abundance),
  nrow = nrow(trichoptera$Abundance),
  ncol = ncol(trichoptera$Abundance)
)
```


The model can be fitted with the function `PLN` as follows:

```{r}
myPLN <- PLN(Abundance ~ 1 + offset(log(TotalCount)), trichoptera)
```

Note the use of the `formula` object to specify the model: the vector $\boldsymbol\mu$ of main effects in the mathematical formulation (one per column species) is specified in the call with the formula `~ 1`. We use the function offset with a log transform of the total count. Abundace is a variable in ther data frame `trichoptera` correspondings to a matrix of 17 columns.

`myPLN` is an `R6` object with class `PLNfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
myPLN
```

Let us try this a bit:

```{r fields-access}
myPLN$loglik
myPLN$BIC
myPLN$ICL
myPLN$R_squared
myPLN$criteria
```

Complementary information comes with the `plot` method:
```{r plot_model, fig.width=7, fig.height=5}
myPLN$plot("model")
```

```{r plot_var, fig.width=7, fig.height=5}
myPLN$plot("variational")
```

