---
title: "Analyzing the Trichopetra data set with the Poisson log-normal LDA model"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLNLDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette intends to illustrate the basical use of the `PLNLDA` function and the methods accompaning the R6 Class `PLNLDAfit`. To this end, let us consider an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence.

```{r data_load}
library(PLNmodels)
data(trichoptera)
grouping <- trichoptera$Group
```

See `?PLNmodels::trichoptera` for details. 

### A discriminant analysis for the Trichopetra data set

We start by adjusting the above model to Trichoptera data set. Here, the matrix of observations $\mathbf{Y}$ corresponds to the abundance of each species in each trapping night. The classification corresponds to the consecutive night groups caught.  It seems natural to add an offset corresponding to the total count per night. The model can be fitted with the function `PLNLDA` as follows:

```{r}
myLDA_nocov <- PLNLDA(Abundance ~ 0 + offset(log(TotalCounts)), grouping = grouping, data = trichoptera)
```

The `myLDA_nocov` variable is an `R6` object with class `PLNLDAfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
myLDA_nocov
```

Essentially, the most important methodis the `plot` method: we separate well the different groups
```{r plot_model, fig.width=7, fig.height=5}
plot(myLDA_nocov)
```

If new samples become available, we can predict their class
```{r predict_class}
predicted.class <- predict(myLDA_nocov, newdata = trichoptera,  type = "response")
```
and asses that we recover the correct class in most (but not all) cases:
```{r check_predicted_class}
table(predicted.class, trichoptera$Group, dnn = c("predicted", "true"))
```

### Accounting for other covariates 

It is possible (experimental) to correct for other covariates before finding the LDA axis that separate well the grouping :
```{r, warning=FALSE}
myLDA_cov <- PLNLDA(Abundance ~ Wind + 0 + offset(log(TotalCounts)), grouping = grouping, data = trichoptera)
```

```{r plot_model_cov, fig.width=7, fig.height=5}
plot(myLDA_cov)
```

And again try to predict class based on the counts and some covariates
```{r predict_class_cov}
predicted.class_cov <- predict(myLDA_cov, newdata = trichoptera, type = "response")
```
and asses that we recover the correct class in most (but not all) cases:
```{r check_predicted_class_cov}
table(predicted.class_cov, trichoptera$Group, dnn = c("predicted", "true"))
```
