---
title: "Analyzing the Trichopetra data set with the Poisson log-normal LDA model"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLNLDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette intends to illustrate the basical use of the `PLNLDA` function and the methods accompaning the R6 Class `PLNLDAfit`. To this end, let us consider an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence.

```{r child = 'trichoptera.Rmd'}
```

### A discriminant analysis for the Trichopetra data set

We start by adjusting the above model to Trichoptera data set. Here, the matrix of observations $\mathbf{Y}$ corresponds to the abundance of each species in each trapping night. The classification corresponds to the consecutive night groups caught. The model can be fitted with the function `PLNLDA` as follows:

```{r}
abundance <- as.matrix(trichometeo$fau) ## must be a matrix
night_grp <- as.factor(trichometeo$cla) 
myLDA_nocov <- PLNLDA(abundance, grouping = night_grp)
```

The `myLDA_nocov` variable is an `R6` object with class `PLNLDAfit`, which comes with a couple of methods, as recalled when printing such an object:

```{r show-method}
myLDA_nocov
```

Essentially, the most important methodis the `plot` method: we separate well the different groups
```{r plot_model, fig.width=7, fig.height=5}
myLDA_nocov$plot_LDA()
```

### Accounting for other covariates 

It is possible (experimental) to correct for other covariates before finding the LDA axis that separate well the grouping :
```{r, warning=FALSE}
abundance <- as.matrix(trichometeo$fau) ## must be a matrix
night_grp <- as.factor(trichometeo$cla) ## must be a factor
covariates <- as.matrix(trichometeo$meteo) ## must be a matrix
myLDA_cov <- PLNLDA(abundance, night_grp, covariates)
```

```{r plot_model_cov, fig.width=7, fig.height=5}
myLDA_cov$plot_LDA()
```
