---
title: "Clustering of multivariate count data with PLN-mixture"
author: "PLN team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: article/PLNreferences.bib
link-citations: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{PLNmixture}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  screenshot.force = FALSE,
  echo = TRUE,
  rows.print = 5,
  message = FALSE,
  warning = FALSE)
```

## Preliminaries

This vignette illustrates the standard use of the `PLNixture` function and the methods accompanying the R6 Classes `PLNmixturefamily` and `PLNmixturefit`.

### Requirements

The packages required for the analysis are **PLNmodels** plus some others for data manipulation and representation:

```{r requirement}
library(PLNmodels)
library(ggplot2)
```

### Data set

We illustrate our point with the trichoptera data set, a full description of which can be found in [the corresponding vignette](Trichoptera.html). Data preparation is also detailed in [the specific vignette](Import_data.html).

```{r data_load}
data(trichoptera)
trichoptera <- prepare_data(trichoptera$Abundance, trichoptera$Covariate)
```

The `trichoptera` data frame stores a matrix of counts (`trichoptera$Abundance`), a matrix of offsets (`trichoptera$Offset`) and some vectors of covariates (`trichoptera$Wind`, `trichoptera$Temperature`, etc.)

### Mathematical background

PLN-mixture for multivariate count data is a variant of the Poisson Lognormal model of @AiH89 (see [the PLN vignette](PLN.html) as a reminder) which can be viewed as a PLN model with an
additional layer to the model by assuming that the latent observations found in the first layer are drawn from a mixture of $K$ multivariate Gaussian components. Each component $k$ have a
prior probability $p(i \in k) = \pi_k$ such that $\sum_k \pi_k = 1$. We denote by $C_i\in \{1,\dots,K\}$ the multinomial variable $\mathcal{M}(1,\matr{\pi} = (\pi_1,\dots,\pi_K))$ describing the
component of observation $i$. Introducing this additional layer, our PLN mixture model is as follows
\begin{equation}
  \begin{array}{rcl}
  \text{layer 2 (clustering)} & \mathbf{C}_i \sim \mathcal{M}(1,\boldsymbol{\pi}) \\
  \text{layer 1 (Gaussian)} &   \mathbf{Z}_i | \, \mathbf{C}_i =  k \sim \mathcal{N}({\boldsymbol\mu}^{(k)}, {\boldsymbol\Sigma}^{(k)}),, \\
  \text{observation space } &  Y_{ij} | Z_{ij} \quad \text{indep.} &   \mathbf{Y}_i | \mathbf{Z}_i\sim\mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right).
  \end{array}
\end{equation}

#### Covariates and offsets

Just like PLN, PLN-mixture generalizes to a formulation where the main effect is due to a linear combination of $d$ covariates $\mathbf{x}_i$ and to a vector $\mathbf{o}_i$ of $p$ offsets in sample $i$ in each mixture component. The latent layer then reads
\begin{equation}
  \mathbf{Z}_i | \mathbf{C}_i = k \, \sim \mathcal{N}({\mathbf{o}_i + \mathbf{x}_i^\top{\boldsymbol\Theta}^{(k)}},{\boldsymbol\Sigma}^{(k)}),
\end{equation}
where each ${\boldsymbol\Theta}^{(k)}$ is a $d\times p$ matrix of regression parameters.

#### Parametrization of the covariance of the mixture models

When using parametric mixture models like Gaussian mixture models, it is generally not recommended to have covariances matrices ${\boldsymbol\Sigma}^{(k)}$ fully parametrized. Here, we implemented either 

  - spherical covariances: ${\boldsymbol\Sigma}^{(k)} =  \left(\sigma^2\right)^{(k)} \mathbf{I}$ 
  - diagonal covariances: ${\boldsymbol\Sigma}^{(k)} =  \mathrm{diag}(\mathbf{d})$ 
  - and fully parametrized covariances.

By default, we use spherical covariances to promote sparsity and for numerical performance. This corresponds to a mixture of Gaussian components with spherical covariance in the latent layer, which is close to a k-means algorithm in the latent Gaussian layer.

#### Optimization by Variational inference

Just like with all models fitted in PLNmodels, we adopt a variational strategy to approximate the log-likelihood function and optimize the consecutive variational surrogate of the log-likelihood with a gradient-ascent-based approach. In this case, it is not too difficult to show that PLN-mixture can be obtained by optimizing a collection of weighted standard PLN model. 

## Analysis of trichoptera data with a PLN-mixture model

In the package, the PLN-mixture model is adjusted with the function `PLNmixture`, which we review in this section. This function adjusts the model for a series of value of $k$ and provides a collection of objects `PLNmixturefit` stored in an object with class `PLNmixturefamily`.

The class `PLNmixturefit` contains a collection of components constituting the mixture, each of whom inherits from the class `PLNfit`, so we strongly recommend the reader to be comfortable with `PLN` and `PLNfit` before using `PLNmixture` (see [the PLN vignette](PLN.html)).

### A mixture model with a latent main effects for the Trichoptera data set

#### Adjusting a collection of fits

We fit a collection of $K$ models as follows:

```{r simple PLNPCA}
mixture_models <- PLNmixture(
  Abundance ~ 1 + offset(log(Offset)),
  data  = trichoptera,
  clusters = 1:5
)
```

Note the use of the `formula` object to specify the model, similar to the one used in the function `PLN`.

#### Structure of `PLNmixturefamily`

The `mixture_models` variable is an `R6` object with class `PLNmixturefamily`, which comes with a couple of methods. The most basic is the `show/print` method, which sends a brief summary of the estimation process:

```{r show nocov}
mixture_models
```

One can also easily access the successive values of the criteria in the collection

```{r collection criteria}
mixture_models$criteria %>% knitr::kable()
```

A quick diagnostic of the optimization process is available via the `convergence` field:

```{r convergence criteria}
mixture_models$convergence  %>% knitr::kable()
```

A visual representation of the optimization can be obtained be representing the objective function
```{r objective}
mixture_models$plot_objective()
```


Comprehensive information about `PLNmixturefamily` is available via `?PLNmixturefamily`.

#### Model selection

The `plot` method of `PLNmixturefamily` displays evolution of the criteria mentioned above, and is a good starting point for model selection:

```{r plot nocov, fig.width=7, fig.height=5}
plot(mixture_models)
```

From this plot, we can see that the best model in terms of BIC or ICL is obtained for a number of cluster `r which.max(mixture_models$BIC)`. We may extract the corresponding model with the method `getBestModel("BIC")`. A model with a specific number of clusters can be extracted with the `getModel()` method:

```{r model extraction}
myMix_BIC <- getBestModel(mixture_models, "BIC")
myMix_3   <- getModel(mixture_models, 3)
```

#### Structure of `PLNmixturefit`

Objects `myMix_ICL` and `myMix_3` are `R6Class` objects of class `PLNmixturefit` which in turns own a couple of methods. A good place to start is the `show/print` method:

```{r map, fig.width=8, fig.height=8}
myMix_BIC
```
```{r}
myMix_BIC$memberships
```

Mixture model can be used to find



```{r plot clustering}
myMix_BIC$plot_clustering_pca()
```

```{r plot reordered data}
myMix_BIC$plot_clustering_data()
```

### A model with latent main effects and meteorological covariates

It is possible to correct for other covariates before finding the LDA axes that best separate well the groups. In our case ,we're going to use `Wind` as a covariate and illustrate the main differences with before  :
```{r, warning=FALSE}
myMix_cov <- PLNmixture(Abundance ~ 1 + Wind + offset(log(Offset)), clusters = 1:3, data = trichoptera)
```


## References

