---
title: "Analyzing the Trichopetra data set with Poisson log-normal Network inference"
author: "Julien Chiquet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
Kvignette: >
  %\VignetteIndexEntry{PLNnetwork}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

This vignette intends to illustrate the basical use of the `PLNnetwork` function and the methods accompaning the R6 Classes `PLNnetworkfamily` and `PLNnetworkfit`. To this end, let us consider an ecological data set concerning abundances of Trichopetera species, coupled with some meteological factors that may influence the presence their presence.

```{r child = 'trichoptera.Rmd'}
```

## Poisson log-normal models

The family of models available in the package belongs to the family of probabilistic PCA models, where the observation are not necessarily Gaussian.  In particular, we consider that observations are drawn from a Poisson distribution.

### A quick view of the mathematical formulation of the model

The Poisson lognormal model relates some $p$-dimensional observation vectors $\mathbf{Y}_i$ to some  $p$-dimensional vectors of Gaussian latent variables $\mathbf{Z}_i$ as follows
\begin{equation} 
  \label{eq:pca-model}
  \begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}{\boldsymbol\mu},\boldsymbol\Sigma) \\
  \text{observation space } &  Y_{ij} | Z_{ij} \quad \text{indep.} & Y_{ij} | Z_{ij} \sim \mathcal{P}\left(\exp\{Z_{ij}\}\right)
  \end{array}
\end{equation}

The parameter ${\boldsymbol\mu}$ corresponds to the main effects and the latent covariance matrix $\boldsymbol\Sigma$ describes the underlying structure of dependence between the $p$ variables. 

In the PLN network, we penalized the inverse covariance $\boldsymbol\Sigma^{-1} = \boldsymbol\Theta$ by a $\ell_1$ penalty to induce sparsity and select important direct relationships between entities. Hence, the support of $\boldsymbol\Theta$ correspond to a network of underlying interactions.

### A model with latent main effects for the Trichopetra data set

We start by adjusting the above model to Trichoptera data set. Here, the matrix of observations $\mathbf{Y}$ corresponds to the abundance of each species in each trapping night. The model can be fitted with the function `PLNNetwork` for a series of penalties contorling the number of edges in the network as follows:

```{r}
abundance <- as.matrix(trichometeo$fau) ## must be a matrix
models <- PLNnetwork(abundance ~ 1)
```

Note the use of the `formula` object to specify the model: the vector $\boldsymbol\mu$ of main effects in the mathematical formulation (one per column species) is specified in the call with the formula `~ 1`.

The `models` variable is an `R6` object with class `PLNnetworkfamily`, which comes with a couple of methods. The most basic is the `show/print` method, which send a very basic summary of the estimation process:

```{r}
models
```
Complementary information comes with the `plot` method:
```{r, fig.width=7, fig.height=5}
models$plot()
```

The critetion $J$ represents the variational lower bound of the likelihood: in this case, we can see that it is strictly increasing with the number of axes (or subspace dimension). Also note the (approximated) $R^2$ which is displayed for each value of penalty. Generally, smoothness of these criteria is an good indicator to check that the optimization process went well.

We can represent the coefficient path (i.e., value of the edge in the network according to the penalty level) to see if some edges clearly come off.

```{r path_coeff, fig.width=7, fig.height=7}
library(ggplot2)
path <- models$coefficient_path(corr = FALSE)
ggplot(path, aes(x = Penalty, y = Coeff, group = Edge, colour = Edge)) + geom_line(show.legend = FALSE) +  coord_trans(x="log10") + theme_bw()
```

An alternative and more intuitive view consiste in plotting the values of the partial correlation along the path.

```{r path_corr, fig.width=7, fig.height=7}
path <- models$coefficient_path(corr = TRUE)
ggplot(path, aes(x = Penalty, y = Coeff, group = Edge, colour = Edge)) + geom_line(show.legend = FALSE) + coord_trans(x="log10") + theme_bw()
```

#### Model selection issue

If we want to select one network, we may extract best model according to BIC or StARS with the method `getBestModel()`. A model with a specific penalty level can be extracted with the `getModel(lambda)` method. 

```{r}
model.BIC   <- models$getBestModel("BIC")   # if no criteria is specified, the best BIC is used
model.StARS <- models$getBestModel("StARS") # if StARS is requested, stabiltiy selection is performed if needed 
model.pen   <- models$getModel(models$penalties[20]) # give some sparsity
```

The variables `model.BIC`, `model.StARS` and `model.pen` are other `R6Class` objects of class `PLNnetworkfit` which in turns owns a couple of methods, mostly for vizualization purposes. The `plot_network` method provides a quick representation of the inferred network, with various options.

```{r support_network, fig.width=7, fig.height=7}
model.BIC$plot_network(type = "support", output = "corrplot")
```

```{r partial_corr_network, fig.width=7, fig.height=7}
par(mfrow = c(1, 3))
model.BIC$plot_network()
model.StARS$plot_network()
model.pen$plot_network()
```

Additional information about the optimization process comes with the `plot_objective` method:
```{r, fig.width=7, fig.height=5}
models$plot_objective()
```
