## The Trichoptera data set

This data set comes with the `R` package *ade4*. We also load our own package for the upcoming analyses:
```{r}
library(PLNmodels)
library(ade4)
data("trichometeo")
```

The insects have been trapped between 1959 and 1960 during 49 trapping nights. These trapping nights are the observations in the case at hand. For each one of these, the abundance of 17 Trichoptera species is recorded, which forms the first table:
```{r}
dim(trichometeo$fau)
head(trichometeo$fau)
```


The second table concerns 11 meteorological variables which may influence the abundance of each species:

```{r}
dim(trichometeo$meteo)
head(trichometeo$meteo)
```
Finally, the observations (that is to say, the trapping nights), have been classified into 12 groups corresponding to contiguous nights between summer 1959 and  summer 1960.

```{r}
barplot(table(trichometeo$cla), main="Trapping nights classificaton")
```

A full description of the data set (in French) is available  [at this page](http://pbil.univ-lyon1.fr/R/pdf/pps034.pdf).
